---
title: Identifying Obstacles to Equitable Flood Mitigation Funding 
author:
  - name: Julian Plough
    affiliations: 
        - ref: unc-eep
  - name: Miyuki Hino
    affiliations:
        - ref: unc-dcrp
        - ref: unc-eep
    orcid: 0000-0001-9369-5769
  - name: Antonia Sebastian
    affiliations:
        - ref: unc-dcrp
        - ref: unc-eep
    orcid: 0000-0002-4309-2561
  - name: Helena Garcia
    affiliations: 
        - ref: unc-eep
    orcid: 0009-0008-3189-1658
  - name: Russell Blessing
    affiliations:
        -  ref: unc-dcrp
    orcid: 0000-0003-4973-4032
affiliations:
- id: unc-dcrp
  name: University of North Carolina at Chapel Hill
  department: Department of Earth, Marine and Environmental Sciences
  city: Chapel Hill
  state: North Carolina
  country: USA
- id: unc-eep
  name: University of North Carolina
  department: Environment, Ecology and Energy Program
  city: Chapel Hill
  state: North Carolina
  country: USA
abstract: |
  As damages from floods, fires, and other climate extremes mount, public investments in risk reduction have also grown. In the US, public spending on risk reduction efforts totaled over $X in 2024, and demand for such funds regularly exceeds available funding. Evidence suggests that in this competitive landscape for funding, urban and whiter communities have fared better at obtaining funding but are often spending it on more disadvantaged neighborhoods within them. However, the reasons for this trend are unclear: complexities in the application process, cost-benefit analysis requirements, and flood damage may all play a role in driving the uneven funding patterns. Here, using a novel dataset tracking properties from flood exposure to application to funding receipt, we identify how each hurdle in the mitigation process shapes the properties that ultimately benefit.  Our findings suggest that application for federal flood mitigation assistance serves as a pinch point where  homeowners in less-white neighborhoods and more affordable homes relative to their neighbors apply. This trend deepens for those who ultimately receive funding, while higher flood-exposure has minimal correlation, implying that funding may not reach all who need it or may be interested. 
keywords: 
  - Flood mitigation
  - Climate resilience
date: last-modified
bibliography: bibliography.bib
format:
  html:
    toc: true
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

# Introduction
Impacts from natural hazards in the United States are increasing in cost and frequency [@noaa2024]. While demand and funding for climate resilience measures has grown significantly in recent years, there are widespread concerns about access to those funds and the extent to which they perpetuate or mitigate existing inequities [@fema2023; @junod2021; @miller2023]. Evidence suggests that while vulnerable groups including mobile home residents and racial minorities are over-represented in flood hotspots, wealthier and whiter communities have received more federal funds to adapt to and recover from floods [@elliot2020; @tate2021]. Several explanations for this disparity have been proposed, including federal cost-benefit analysis requirements, limited local government capacity to apply for grants, and lack of household awareness about funding opportunities. However, research to pinpoint funding obstacles has been constrained by only observing who ultimately receives funding, not who is eligible or who applies for funding.As a result, it is unclear what the barriers are, and therefore which policy or procedural changes would most improve access to climate resilience funds.

Here, we combine novel datasets on over two decades of flood exposure and applications for federal funding in North Carolina to examine the full process: who floods, who applies for mitigation funding, and who receives it. This flooded-to-funded pipeline (Fig. 1) is critical for understanding the process by which households experience the Hazard Mitigation Assistance (HMA) process. By analyzing the household- and neighborhood-level characteristics at each stage of the process, we provide new evidence regarding how the application process shapes the distribution of funding. For example, if households located in marginalized communities rarely submit applications, the primary issue may be lack of awareness or resources for applying. In contrast, if they are applying but not being funded, the selection criteria may be driving disproportionate access.

Our initial results demonstrate that the application stage is crucial for understanding the mitigation funding pipeline. 67.5% of properties that experience flooding are eligible for public assistance, and 61% are in communities where local governments have submitted applications for assistance. However, only 2.3% of all flooded properties apply and ~1% are funded. Properties that apply are significantly more likely to have experienced more than one flooding event over the period. Compared to the population of flooded properties, properties that apply for funding have lower property values (58th percentile to 42nd percentile) and come from census block groups with a lower share of white residents (74% to 64%). Among properties receiving funding, property values and the share of white residents in the block group are even lower. These results are consistent with some nationwide findings, where wealthier and whiter communities have been linked to mitigation access [mach2019; Elliot 2020], however we add fidelity in that applying and funded properties are also lower-in-value than their flooded neighbors. Notably, these trends emerge at the newly-visible application stage and widen for those ultimately funded. We are the first to demonstrate that unpacking and examining this stage is pivotal to understanding potential obstacles to equitable mitigation funding.

# Bibliography styles



With this template using elsevier class, natbib will be used. Three bibliographic style ﬁles (*.bst) are provided and their use controled by `cite-style` option: 

- `citestyle: number` (default)  will use `elsarticle-num.bst` - can be used for the numbered scheme
- `citestyle: numbername` will use `elsarticle-num-names.bst` - can be used for numbered with new options of natbib.sty
- `citestyle: authoryear` will use `elsarticle-harv.bst` — can be used for author year scheme

This `citestyle` will insert the right `.bst` and set the correct `classoption` for `elsarticle` document class.

Using `natbiboptions` variable in YAML header, you can set more options for `natbib` itself . Example 

```yaml
natbiboptions: longnamesfirst,angle,semicolon
```

## Using CSL 

If `cite-method` is set to `citeproc` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. By default, this template will provide an appropriate style, but alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded
and stored locally, or the url can be used as in the example header.

# Equations

Here is an equation:
$$ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
$$

Inline equations work as well: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables

@fig-meaningless is generated using an R chunk.

```{r}
#| label: fig-meaningless
#| fig-cap: A meaningless scatterplot
#| fig-width: 5
#| fig-height: 5
#| fig-align: center
#| out-width: 50%
#| echo: false
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in @tbl-simple example.

```{r}
#| label: tbl-simple
#| tbl-cap: Caption centered above table
#| echo: true
knitr::kable(head(mtcars)[,1:4])
```

# References {-}
