---
title: Identifying Obstacles to Equitable Flood Mitigation Funding 
author:
  - name: Julian Plough
    affiliations: 
        - id: unc-eep
  - name: Miyuki Hino
    affiliations:
        - unc-dcrp
        - unc-eep
    orcid: 0000-0001-9369-5769
  - name: Antonia Sebastian
    affiliations:
        - unc-dcrp
        - unc-eep
    orcid: 0000-0002-4309-2561
  - name: Helena Garcia
    affiliations: 
        - id: unc-eep
    orcid: 0009-0008-3189-1658
  - name: Russell Blessing
    affiliations:
        - id: unc-dcrp
    orcid: 0000-0003-4973-4032
affiliationns:
- id: unc-dcrp
  name: University of North Carolina at Chapel Hill
  department: Department of Earth, Marine and Environmental Sciences
  city: Chapel Hill
  state: North Carolina
  country: USA
- id: unc-eep
  name: University of North Carolina
  department: Environment, Ecology and Energy Program
  city: Chapel Hill
  state: North Carolina
  country: USA
abstract: |
As damages from floods, fires, and other climate extremes mount, public investments in risk reduction have also grown. In the US, public spending on risk reduction efforts totaled over $X in 2024, and demand for such funds regularly exceeds available funding. Evidence suggests that in this competitive landscape for funding, urban and whiter communities have fared better at obtaining funding but are often spending it on more disadvantaged neighborhoods within them. However, the reasons for this trend are unclear: complexities in the application process, cost-benefit analysis requirements, and flood damage may all play a role in driving the uneven funding patterns. Here, using a novel dataset tracking properties from flood exposure to application to funding receipt, we identify how each hurdle in the mitigation process shapes the properties that ultimately benefit.  Our findings suggest that application for federal flood mitigation assistance serves as a pinch point where  homeowners in less-white neighborhoods and more affordable homes relative to their neighbors apply. This trend deepens for those who ultimately receive funding, while higher flood-exposure has minimal correlation, implying that funding may not reach all who need it or may be interested. 
keywords: 
  - Flood mitigation
  - Climate resilience
date: last-modified
bibliography: bibliography.bib
format:
  html:
    toc: true
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      # model: 3p # Don't set a model with preprint
      cite-style: authoryear
---

Please make sure that your manuscript follows the guidelines in the 
Guide for Authors of the relevant journal. It is not necessary to 
typeset your manuscript in exactly the same way as an article, 
unless you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>

# Bibliography styles

Here are two sample references:  @Feynman1963118 @Dirac1953888.

With this template using elsevier class, natbib will be used. Three bibliographic style ﬁles (*.bst) are provided and their use controled by `cite-style` option: 

- `citestyle: number` (default)  will use `elsarticle-num.bst` - can be used for the numbered scheme
- `citestyle: numbername` will use `elsarticle-num-names.bst` - can be used for numbered with new options of natbib.sty
- `citestyle: authoryear` will use `elsarticle-harv.bst` — can be used for author year scheme

This `citestyle` will insert the right `.bst` and set the correct `classoption` for `elsarticle` document class.

Using `natbiboptions` variable in YAML header, you can set more options for `natbib` itself . Example 

```yaml
natbiboptions: longnamesfirst,angle,semicolon
```

## Using CSL 

If `cite-method` is set to `citeproc` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. By default, this template will provide an appropriate style, but alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded
and stored locally, or the url can be used as in the example header.

# Equations

Here is an equation:
$$ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
$$

Inline equations work as well: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables

@fig-meaningless is generated using an R chunk.

```{r}
#| label: fig-meaningless
#| fig-cap: A meaningless scatterplot
#| fig-width: 5
#| fig-height: 5
#| fig-align: center
#| out-width: 50%
#| echo: false
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in @tbl-simple example.

```{r}
#| label: tbl-simple
#| tbl-cap: Caption centered above table
#| echo: true
knitr::kable(head(mtcars)[,1:4])
```

# References {-}
