---
title: "Step 6: Property Value Interpolation & EIF Census Enrichment"
author: "Russell Blessing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show / hide code"
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This notebook does two related things:

**Property value interpolation.** For parcels matched to a mitigation or
application record but carrying an unreliable `TOTAL.VALUE.CALCULATED`
(vacant land, government-owned, anomalously low values), estimate a
market-comparable value by averaging the recorded values of the nearest
1–4 family residential donor parcels. Three estimates per recipient
(k = 5, 10, 50 nearest neighbors).

**EIF census enrichment.** For each mit/app point, attach Census Bureau
Gridded EIF attributes from the 2024 release at the 0.01° grid cell
containing that point. Three EIF files joined: housing units / age /
home value, population × age × race × sex, and population × race ×
income decile.

Both stages consume the **alt Step 5 resolved outputs** as the single source of
mit/app data. Each record carries an `assigned_parcel_index` (resolved
to one parcel per record, no fan-out) and a point geometry. The
interpolation uses `assigned_parcel_index` to count mits/apps per
parcel; the EIF enrichment uses the point geometry to snap each record
to its grid cell.

**Inputs:**

- `parcels_pri.gpkg` — parcels with priority scoring (canonical Step 5)
- `alt_mits_resolved.gpkg` — point-geometry mits, one row per record,
  resolved to one assigned parcel each (alt Step 5)
- `alt_apps_resolved.gpkg` — point-geometry apps, same structure (alt Step 5)
- Three Gridded EIF parquet files (downloaded on first run)

**Outputs (written to `out_dir`):**

- `interpolate_to.gpkg`, `interpolate_from.gpkg` — recipient and donor pools
- `interpolated_vals.gpkg`, `interpolated_parcels_final.csv` — parcels with interpolated values
- `mits_with_eif.gpkg`, `apps_with_eif.gpkg` — alt resolved outputs with EIF cell attributes
- `mits_with_eif.csv`, `apps_with_eif.csv` — same, no geometry

## Methodology

**Recipients (parcels with unreliable recorded property values)**: parcels
that have at least one mit or app assigned to them by alt Step 5 AND
whose recorded `TOTAL.VALUE.CALCULATED` is treated as unreliable.
A value is treated as unreliable if the parcel is:

- vacant land (`vacant == 1`) — no structure, recorded value reflects land only
- government land use (`gov_lu == 1`) — typically tax-exempt with $0 assessed
- government-owned by owner name (`gov_owned == 1`) — same
- low recorded value (`TOTAL.VALUE.CALCULATED < $25,000`) — anomalously low,
  likely indicates undervaluation, recent transfer, or data quality issue

**Donors (parcels providing value estimates)**: 1–4 family residential,
NOT government-owned, with non-missing recorded total value above
$25,000. Strict NA exclusion ensures every neighbor mean is computed
over exactly k actual values.

**Mit/app per-parcel counts** are computed from `assigned_parcel_index`
in the alt resolved outputs — each unique mit/app contributes to
exactly one parcel. This avoids the fan-out double-counting that would
happen if we counted from canonical `mits_pcls.gpkg` (where a mit
hitting N parcels appears in N rows). Records with `assignment_method
== "unassigned"` are excluded from counts.

**Government-ownership detection** uses a regex on `OWNER.1.FULL.NAME`
matching common government-entity keywords (`CITY`, `STATE`, `TOWN`,
`VILLAGE`, `COUNTY`, `METROPOLITAN`, plus abbreviations) excluding
`INC` and `LLC`.

**EIF cell join** uses cell-membership rather than spatial intersection —
each mit/app point's lat/lon is snapped to its 0.01° grid cell corner
via `floor(coord * 100) / 100`, then joined to EIF data by the snapped
coordinates as a key. Equivalent result, much faster than `st_join`.

**Caveats** worth noting in the manuscript:

- The $25,000 thresholds are heuristic; sensitivity analysis at $10k
  and $50k recommended.
- Three K values (5/10/50) are produced in parallel; downstream
  analysis picks the granularity.
- EIF counts are differentially-private — approximate per cell.
- 1.2 km² EIF cells share between adjacent condo buildings.
- Donor pool is statewide; border parcels may reach across counties.
- Government detection uses owner-name regex (or CoreLogic exemption
  code if present — see Section 3).

```{r libraries}
library(sf)
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(purrr)
library(RANN)
library(arrow)
library(knitr)
```

```{r setup}
out_dir         <- "/proj/mhinolab/users/rbless/data/Obstacles_Output"
parcels_path    <- file.path(out_dir, "parcels_pri.gpkg")
mits_alt_path   <- file.path(out_dir, "alt_mits_resolved.gpkg")
apps_alt_path   <- file.path(out_dir, "alt_apps_resolved.gpkg")

# Tunables — interpolation
LOW_VALUE_THRESHOLD <- 25000   # parcels with TOTAL.VALUE.CALCULATED below this
                                # are flagged as needing interpolation
MIN_DONOR_VALUE     <- 25000   # minimum recorded value for a donor parcel
K_VALUES            <- c(5, 10, 50)

# Tunables — government ownership regex
GOV_KEYWORDS     <- "\\bCITY\\b|\\bSTATE\\b|\\bTOWN\\b|\\bVILLAGE\\b|\\bCOUNTY\\b|\\bMETROPOLITAN\\b|\\bCTY\\b|\\bST\\b|\\bTWN\\b|\\bVILL\\b|\\bVIL\\b|\\bCNTY\\b|\\bMETRO\\b"
EXCLUDE_KEYWORDS <- "\\bINC\\b|\\bLLC\\b"

# EIF source files (2024 release)
eif_files <- list(
  hu_age_homeval = list(
    url    = "https://www2.census.gov/ces/gridded_eif/gridded_eif_hu_age_homeval_2024.parquet",
    path   = file.path(out_dir, "gridded_eif_hu_age_homeval_2024.parquet"),
    prefix = "hu"
  ),
  ageracesex = list(
    url    = "https://www2.census.gov/ces/gridded_eif/gridded_eif_pop_ageracesex_2024.parquet",
    path   = file.path(out_dir, "gridded_eif_pop_ageracesex_2024.parquet"),
    prefix = "ars"
  ),
  raceincome = list(
    url    = "https://www2.census.gov/ces/gridded_eif/gridded_eif_pop_raceincome_2024.parquet",
    path   = file.path(out_dir, "gridded_eif_pop_raceincome_2024.parquet"),
    prefix = "ri"
  )
)

# NC bounding box (WGS84) with small buffer for cells crossing the boundary
NC_BBOX <- list(lat_min = 33.7, lat_max = 36.7,
                lon_min = -84.5, lon_max = -75.4)
```

## 1  Load parcels and alt mits/apps

```{r load-parcels}
parcels <- st_read(parcels_path, quiet = TRUE) |>
  st_transform(32119)

cat("Parcels loaded:", nrow(parcels), "rows\n")
```

```{r load-mits-apps}
mits_resolved <- st_read(mits_alt_path, quiet = TRUE)
apps_resolved <- st_read(apps_alt_path, quiet = TRUE)

cat("Mits (alt resolved):", nrow(mits_resolved), "rows\n")
cat("Apps (alt resolved):", nrow(apps_resolved), "rows\n")

# Quick distribution of how each side was assigned
cat("\nMit assignment_method distribution:\n")
mits_resolved |> st_drop_geometry() |> count(assignment_method, sort = TRUE) |> print()
cat("\nApp assignment_method distribution:\n")
apps_resolved |> st_drop_geometry() |> count(assignment_method, sort = TRUE) |> print()
```

## 2  Aggregate mit/app counts per parcel

One mit/app contributes to exactly one parcel — counted by
`assigned_parcel_index` from the alt resolved outputs. Records with
`assignment_method == "unassigned"` (no parcel within reach) are
excluded. Parcels with no matched mits/apps get a zero count.

```{r mit-app-counts}
mit_counts <- mits_resolved |>
  st_drop_geometry() |>
  filter(!is.na(assigned_parcel_index)) |>
  count(assigned_parcel_index, name = "mit_count") |>
  rename(parcel_index = assigned_parcel_index)

app_counts <- apps_resolved |>
  st_drop_geometry() |>
  filter(!is.na(assigned_parcel_index)) |>
  count(assigned_parcel_index, name = "app_count") |>
  rename(parcel_index = assigned_parcel_index)

parcels <- parcels |>
  left_join(mit_counts, by = "parcel_index") |>
  left_join(app_counts, by = "parcel_index") |>
  mutate(
    mit_count = replace_na(mit_count, 0L),
    app_count = replace_na(app_count, 0L)
  )

cat("Parcels with at least one assigned mit:", sum(parcels$mit_count > 0), "\n")
cat("Parcels with at least one assigned app:", sum(parcels$app_count > 0), "\n")
cat("Parcels with both:                     ",
    sum(parcels$mit_count > 0 & parcels$app_count > 0), "\n")
```

## 3  Identify government-owned parcels

```{r gov-owned}
parcels <- parcels |>
  mutate(
    owner_name_clean = replace_na(`OWNER.1.FULL.NAME`, ""),
    gov_owned = as.integer(
      str_detect(owner_name_clean, regex(GOV_KEYWORDS,    ignore_case = TRUE)) &
      !str_detect(owner_name_clean, regex(EXCLUDE_KEYWORDS, ignore_case = TRUE))
    )
  ) |>
  select(-owner_name_clean)

cat("Government-owned parcels (by owner name):", sum(parcels$gov_owned), "\n")
cat("Overlap with gov_lu (by land use code):  ",
    sum(parcels$gov_owned == 1 & parcels$gov_lu == 1), "\n")
```

## 4  Type conversions

`TOTAL.VALUE.CALCULATED` and `IMPROVEMENT.VALUE.CALCULATED` came in as
character. Coerce to numeric for the threshold filters and KNN means.

```{r type-conversions}
parcels <- parcels |>
  mutate(
    `TOTAL.VALUE.CALCULATED`       = suppressWarnings(as.numeric(`TOTAL.VALUE.CALCULATED`)),
    `IMPROVEMENT.VALUE.CALCULATED` = suppressWarnings(as.numeric(`IMPROVEMENT.VALUE.CALCULATED`)),
    fr_1_4 = as.integer(fr_1_4)
  )
```

## 5  Define recipient and donor pools

```{r build-pools}
to_interpolate <- parcels |>
  filter(
    (mit_count > 0 | app_count > 0),
    (
      vacant == 1 |
      gov_lu == 1 |
      gov_owned == 1 |
      (!is.na(`TOTAL.VALUE.CALCULATED`) & `TOTAL.VALUE.CALCULATED` < LOW_VALUE_THRESHOLD)
    )
  )

interpolate_from <- parcels |>
  filter(
    fr_1_4 == 1,
    gov_owned == 0,
    !is.na(`TOTAL.VALUE.CALCULATED`),
    `TOTAL.VALUE.CALCULATED` > MIN_DONOR_VALUE
  )

cat("Recipients (to_interpolate):", nrow(to_interpolate), "\n")
cat("Donors (interpolate_from):  ", nrow(interpolate_from), "\n")

st_write(to_interpolate,   file.path(out_dir, "interpolate_to.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
st_write(interpolate_from, file.path(out_dir, "interpolate_from.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
```

## 6  K-nearest-neighbor interpolation

For each recipient parcel, find the K nearest donor parcels by centroid
distance, then average their `TOTAL.VALUE.CALCULATED`. Three values of
K (5, 10, 50) computed in parallel.

```{r knn-interpolation}
from_coords <- st_coordinates(st_centroid(interpolate_from))
to_coords   <- st_coordinates(st_centroid(to_interpolate))

donor_values <- interpolate_from$`TOTAL.VALUE.CALCULATED`

nn_results <- map(K_VALUES, ~ RANN::nn2(from_coords, to_coords, k = .x))
names(nn_results) <- paste0("k", K_VALUES)

prop_values <- map2_dfc(K_VALUES, nn_results, function(k, nn) {
  vals <- matrix(donor_values[nn$nn.idx], ncol = k)
  tibble(!!paste0("prop_value_", k) := rowMeans(vals))
})

to_interpolate <- bind_cols(to_interpolate, prop_values)

cat("Recipient row count:", nrow(to_interpolate), "\n")
cat("All prop_value_5 non-NA:",  all(!is.na(to_interpolate$prop_value_5)),  "\n")
cat("All prop_value_10 non-NA:", all(!is.na(to_interpolate$prop_value_10)), "\n")
cat("All prop_value_50 non-NA:", all(!is.na(to_interpolate$prop_value_50)), "\n")
```

## 7  Join interpolated values back to all parcels

```{r join-back}
to_interp_attrs <- to_interpolate |>
  st_drop_geometry() |>
  select(parcel_index, prop_value_5, prop_value_10, prop_value_50)

joined <- parcels |>
  left_join(to_interp_attrs, by = "parcel_index") |>
  mutate(
    interpolate = as.integer(parcel_index %in% to_interpolate$parcel_index),
    prop_value_5  = coalesce(prop_value_5,  `TOTAL.VALUE.CALCULATED`),
    prop_value_10 = coalesce(prop_value_10, `TOTAL.VALUE.CALCULATED`),
    prop_value_50 = coalesce(prop_value_50, `TOTAL.VALUE.CALCULATED`)
  )

cat("Total parcels in joined output:", nrow(joined), "\n")
cat("Parcels marked as interpolated:", sum(joined$interpolate == 1), "\n")
```

## 8  Write interpolated parcels output

```{r write-interpolation-outputs}
st_write(joined, file.path(out_dir, "interpolated_vals.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

joined |>
  st_drop_geometry() |>
  write_csv(file.path(out_dir, "interpolated_parcels_final.csv"))
```

## 9  Gridded EIF Census enrichment

The Census Bureau's 2024 Gridded EIF release provides privacy-protected
counts on a fixed 0.01° grid (~1.2 km² per cell). Three datasets joined
to each mit/app point:

| File | Cross-tab | Variable prefix |
|---|---|---|
| `gridded_eif_hu_age_homeval_2024.parquet` | housing units × householder age × income × home value | `hu_` |
| `gridded_eif_pop_ageracesex_2024.parquet` | population × age × race × sex | `ars_` |
| `gridded_eif_pop_raceincome_2024.parquet` | population × race × income decile | `ri_` |

Cell-membership join via `floor(coord * 100) / 100` — equivalent to
spatial intersection but faster.

### 9.1  Download EIF files (one-time)

```{r eif-download}
for (name in names(eif_files)) {
  f <- eif_files[[name]]
  if (!file.exists(f$path)) {
    message("Downloading ", name, " -> ", f$path)
    download.file(f$url, f$path, mode = "wb")
  } else {
    message("Already present: ", basename(f$path))
  }
}
```

### 9.2  Inspect EIF schemas

Print the column names from each parquet so the join configuration below
can reference real column names. Update `LAT_COL` / `LON_COL` if the
schemas use different names than `lat` / `lon`.

```{r eif-inspect-schemas}
for (name in names(eif_files)) {
  f <- eif_files[[name]]
  cat("\n=== ", name, " ===\n")
  print(arrow::open_dataset(f$path)$schema)
}
```

### 9.3  Read EIF parquets (filtered to NC bounding box)

```{r eif-read}
LAT_COL <- "grid_lat"      # confirmed via topology — likely the column name in parquet too
LON_COL <- "grid_lon"      # confirm via eif-inspect-schemas

read_eif <- function(path, prefix) {
  d <- arrow::open_dataset(path) |>
    filter(
      .data[[LAT_COL]] >= NC_BBOX$lat_min,
      .data[[LAT_COL]] <= NC_BBOX$lat_max,
      .data[[LON_COL]] >= NC_BBOX$lon_min,
      .data[[LON_COL]] <= NC_BBOX$lon_max
    ) |>
    collect()

  # Cell key columns — use parquet's stored center coords directly,
  # rounded to 3 decimals to avoid float comparison issues on join
  d <- d |>
    mutate(
      eif_lat = round(.data[[LAT_COL]], 3),
      eif_lon = round(.data[[LON_COL]], 3)
    )

  data_cols <- setdiff(names(d), c("eif_lat", "eif_lon"))
  rename_map <- setNames(data_cols, paste0(prefix, "_", data_cols))
  d |> rename(!!!rename_map)
}

eif_hu  <- read_eif(eif_files$hu_age_homeval$path, eif_files$hu_age_homeval$prefix)
eif_ars <- read_eif(eif_files$ageracesex$path,     eif_files$ageracesex$prefix)
eif_ri  <- read_eif(eif_files$raceincome$path,     eif_files$raceincome$prefix)

cat("EIF cells in NC bounding box:\n")
cat("  hu_age_homeval:", nrow(eif_hu),  "\n")
cat("  ageracesex:    ", nrow(eif_ars), "\n")
cat("  raceincome:    ", nrow(eif_ri),  "\n")
```

### 9.4  Snap mit/app points to EIF cells

`mits_resolved` and `apps_resolved` were already loaded in Section 1.
Compute the snapped cell key for each.

```{r eif-snap-cells}
add_eif_cell <- function(sf_obj) {
  coords <- sf_obj |> st_transform(4326) |> st_coordinates()
  sf_obj |>
    mutate(
      # 0.01° cell, centers at .005 offsets per eif_grid_topology
      eif_lon = round(floor(coords[, "X"] * 100) / 100 + 0.005, 3),
      eif_lat = round(floor(coords[, "Y"] * 100) / 100 + 0.005, 3)
    )
}

mits_resolved <- add_eif_cell(mits_resolved)
apps_resolved <- add_eif_cell(apps_resolved)
```

### 9.5  Join all three EIF datasets onto mits/apps

```{r eif-join}
join_eif_all <- function(sf_obj) {
  sf_obj |>
    left_join(eif_hu,  by = c("eif_lat", "eif_lon")) |>
    left_join(eif_ars, by = c("eif_lat", "eif_lon")) |>
    left_join(eif_ri,  by = c("eif_lat", "eif_lon"))
}

mits_with_eif <- join_eif_all(mits_resolved)
apps_with_eif <- join_eif_all(apps_resolved)

# Coverage diagnostics — % of records that got EIF data from each file.
hu_marker  <- paste0("hu_",  LAT_COL)
ars_marker <- paste0("ars_", LAT_COL)
ri_marker  <- paste0("ri_",  LAT_COL)

coverage <- function(df, label) {
  cat("\n", label, "EIF coverage (cells matched / total records):\n")
  cat("  hu_age_homeval:",
      sprintf("%d / %d (%.1f%%)\n",
              sum(!is.na(df[[hu_marker]])),  nrow(df),
              100 * sum(!is.na(df[[hu_marker]]))  / nrow(df)))
  cat("  ageracesex:    ",
      sprintf("%d / %d (%.1f%%)\n",
              sum(!is.na(df[[ars_marker]])), nrow(df),
              100 * sum(!is.na(df[[ars_marker]])) / nrow(df)))
  cat("  raceincome:    ",
      sprintf("%d / %d (%.1f%%)\n",
              sum(!is.na(df[[ri_marker]])),  nrow(df),
              100 * sum(!is.na(df[[ri_marker]]))  / nrow(df)))
}

coverage(mits_with_eif, "Mitigations")
coverage(apps_with_eif, "Applications")
```

### 9.6  Write enriched outputs

```{r eif-write}
st_write(mits_with_eif, file.path(out_dir, "mits_with_eif.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
st_write(apps_with_eif, file.path(out_dir, "apps_with_eif.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

mits_with_eif |> st_drop_geometry() |>
  write_csv(file.path(out_dir, "mits_with_eif.csv"))
apps_with_eif |> st_drop_geometry() |>
  write_csv(file.path(out_dir, "apps_with_eif.csv"))
```

## Summary

```{r summary-table}
summary_tbl <- tibble(
  metric = c(
    "Total parcels",
    "Recipients (interpolated)",
    "Donors (used for averaging)",
    "Median original TOTAL.VALUE.CALCULATED (recipients only)",
    "Median prop_value_5",
    "Median prop_value_10",
    "Median prop_value_50",
    "Mits enriched with EIF (hu_age_homeval cell match)",
    "Apps enriched with EIF (hu_age_homeval cell match)"
  ),
  value = c(
    nrow(joined),
    sum(joined$interpolate == 1),
    nrow(interpolate_from),
    median(to_interpolate$`TOTAL.VALUE.CALCULATED`, na.rm = TRUE),
    median(to_interpolate$prop_value_5,  na.rm = TRUE),
    median(to_interpolate$prop_value_10, na.rm = TRUE),
    median(to_interpolate$prop_value_50, na.rm = TRUE),
    sum(!is.na(mits_with_eif[[hu_marker]])),
    sum(!is.na(apps_with_eif[[hu_marker]]))
  )
)

knitr::kable(summary_tbl, format.args = list(big.mark = ","))
```

```{r value-comparison}
to_interpolate |>
  st_drop_geometry() |>
  select(parcel_index, original = `TOTAL.VALUE.CALCULATED`,
         prop_value_5, prop_value_10, prop_value_50) |>
  pivot_longer(-parcel_index, names_to = "metric", values_to = "value") |>
  group_by(metric) |>
  summarise(
    n      = sum(!is.na(value)),
    p10    = quantile(value, 0.10, na.rm = TRUE),
    median = median(value,           na.rm = TRUE),
    mean   = mean(value,             na.rm = TRUE),
    p90    = quantile(value, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(
    format.args = list(big.mark = ",", scientific = FALSE),
    caption = "Distribution of original vs. interpolated values for recipient parcels"
  )
```

## Output schema

**`interpolated_vals.gpkg`** / **`interpolated_parcels_final.csv`** — one row per parcel:

| Column | Source | Notes |
|---|---|---|
| (all original parcel columns) | parcels_pri.gpkg | unchanged |
| `mit_count` | derived | per-parcel count of mits assigned via alt Step 5 |
| `app_count` | derived | per-parcel count of apps assigned via alt Step 5 |
| `gov_owned` | derived | 1 if owner name matches gov regex (excluding INC/LLC) |
| `interpolate` | derived | 1 if parcel was a recipient (had value estimated) |
| `prop_value_5` | derived | mean `TOTAL.VALUE.CALCULATED` of 5 nearest donors; original if not interpolated |
| `prop_value_10` | derived | mean of 10 nearest donors; original if not interpolated |
| `prop_value_50` | derived | mean of 50 nearest donors; original if not interpolated |

**`mits_with_eif.gpkg`** / **`apps_with_eif.gpkg`** — one row per mit/app:

| Column | Source | Notes |
|---|---|---|
| (all original alt-resolved columns) | alt Step 5 outputs | unchanged |
| `eif_lat_floor`, `eif_lon_floor` | derived | snapped 0.01° cell key |
| `hu_*` columns | EIF hu_age_homeval | housing units, householder age, home value |
| `ars_*` columns | EIF pop_ageracesex | population × age × race × sex counts |
| `ri_*` columns | EIF pop_raceincome | population × race × income decile counts |

## Unresolved — confirm before using outputs

- **`LOW_VALUE_THRESHOLD = 25000`** and **`MIN_DONOR_VALUE = 25000`** are
  heuristic. Sensitivity analysis at $10k and $50k recommended.
- **`K_VALUES = c(5, 10, 50)`** — three parallel estimates. Drop unused
  K values once a preferred granularity is selected.
- **No within-county constraint** on donor matching. Border parcels may
  pull donors across county lines.
- **Government-ownership regex** is empirically derived. Spot-check
  `gov_owned == 1` against `parusedesc` to confirm coverage.
- **EIF `LAT_COL` / `LON_COL` names** assumed to be `lat` / `lon`.
  Confirm via `eif-inspect-schemas` chunk and update if different.
- **EIF year choice** — using 2024 release for all mits/apps regardless
  of program year. Historical EIF files (1999–2023) are available if
  you want time-matched demographics for older program records.
  Defensible methodological choice (current demographics as a baseline)
  but worth flagging in the methods note.
- **EIF differential privacy** — per-cell counts are noise-protected.
  Negligible for averages across many cells, important for per-cell
  point estimates.
- **`unassigned` mits/apps** are excluded from per-parcel counts but
  still appear in the EIF-enriched outputs (with NA `assigned_parcel_index`).
  Downstream analysis should decide whether to include or exclude these
  records depending on whether the analytic unit is the mit/app record
  or the matched parcel.