---
title: "Step 5.1 (Alt): Direct Intersection — Mits & Apps to Parcels"
author: "Russell Blessing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show / hide code"
execute:
  echo: true
  warning: false
  message: false
---

## Overview

A transparent, no-heuristic spatial join of mitigation and application records
to parcels. Every input record is kept (no deduplication). A single point
representing, say, six mitigations on a condo building stays as six rows in
the output. Shared-footprint parcels fan out naturally: one input record that
intersects *N* parcel polygons produces *N* output rows.

This file is **complementary** to `05_spatial_match_app_mit.qmd`, which runs
the tiered priority cascade and produces the canonical matched output. Nothing
here changes that output.

**What this file does NOT do:**

- No tier cascade (high → med → low → fallback NN).
- No distance-based matching or buffering in the join itself.
- No dedup of mit/app records.
- No winner selection when a point hits multiple shared-footprint parcels.

**Inputs:**

- `parcels_pri.gpkg` — parcels with `luc_tier` column (from `out_dir`)
- `mit_newgeo.gpkg` — raw mitigations (`CRC_May_2025/Raw Data Inputs/raw_app_ncb/`)
- `applications_final.gpkg` — raw applications (same directory)

**Outputs (written to `out_dir`):**

- `alt_mits_direct.gpkg`, `alt_apps_direct.gpkg` — full fan-out spatial join
- `alt_mits_direct_review.csv`, `alt_apps_direct_review.csv` — same, no geometry

---

```{r libraries}
library(sf)
library(dplyr)
library(readr)
library(tidyr)
library(purrr)
library(knitr)
library(DT)
library(leaflet)
library(glue)
```

```{r setup}
out_dir <- "/proj/mhinolab/users/rbless/data/Obstacles_Output"

parcels_path <- file.path(out_dir, "parcels_pri.gpkg")
mits_path    <- "/proj/mhinolab/projects/obstacles2/CRC_May_2025/Raw Data Inputs/raw_app_ncb/mit_newgeo.gpkg"
apps_path    <- "/proj/mhinolab/projects/obstacles2/CRC_May_2025/Raw Data Inputs/raw_app_ncb/applications_final.gpkg"

# luc_tier >= HIGH_TIER_MIN is considered "high priority" for tagging and
# proximity review. Tiers 3 and 4 = vacant/gov/residential + 1-4 family.
# Flip to 4L if supervisors want tier-4-only (1-4 family + site address only).
HIGH_TIER_MIN <- 3L

# Primary-key columns on the mit/app sides.
# Both default to "unique_id", assigned by row_number() at load time below —
# keeps the script independent of canonical Step 5.
# Adjust if the raw inputs use a different HMA/IHP key.
mit_id_col <- "unique_id"
app_id_col <- "unique_id"
```

---

## 1  Load inputs

```{r load-inputs}
parcels <- st_read(parcels_path, quiet = TRUE) |>
  select(
    parcel_index, geometry_id, cntyfips, parno, altparno,
    siteadd, scity, szip, mailadd, ownname, parusedesc,
    `LAND.USE.CODE`, luc_tier,
    `SITUS.HOUSE.NUMBER`, `SITUS.STREET.NAME`,
    `OWNER.1.FULL.NAME`, `YEAR.BUILT`, `ASSESSED.TOTAL.VALUE`
  ) |>
  st_transform(32119)

# Do NOT dedup — every HMA / IHP record is preserved, coordinate duplicates
# included. A point shared by six mitigations stays as six rows.
mits <- st_read(mits_path, quiet = TRUE) |>
  st_transform(st_crs(parcels)) |>
  mutate(unique_id = row_number())
apps <- st_read(apps_path, quiet = TRUE) |>
  st_transform(st_crs(parcels)) |>
  mutate(unique_id = row_number())
```

---

## 2  Direct intersection join

One row per (input record × intersecting parcel). Records that fall outside
every parcel produce one row with NA parcel columns (`left = TRUE`).

```{r direct-join}
mits_direct <- st_join(mits, parcels, join = st_intersects, left = TRUE)
apps_direct <- st_join(apps, parcels, join = st_intersects, left = TRUE)
```

---

## 3  Tag match status

```{r tag-status}
tag_status <- function(df, high_tier_min = HIGH_TIER_MIN) {
  df |>
    mutate(
      luc_tier_num = suppressWarnings(as.integer(luc_tier)),
      match_status = case_when(
        is.na(parcel_index)           ~ "no_parcel_hit",
        luc_tier_num >= high_tier_min ~ "matched_high",
        TRUE                          ~ "matched_low"
      )
    )
}

mits_direct <- tag_status(mits_direct)
apps_direct <- tag_status(apps_direct)
```

---

## 4  Proximity to nearest high-priority parcel

For every input point, compute distance to the nearest high-priority parcel.
This is for **review only** — it does not change any match result.

```{r proximity-to-high}
high_parcels <- parcels |> filter(as.integer(luc_tier) >= HIGH_TIER_MIN)

nearest_high <- function(pts, id_col, high) {
  if (nrow(pts) == 0 || nrow(high) == 0) {
    return(tibble(
      !!id_col := character(),
      nearest_high_parcel_index = integer(),
      dist_to_nearest_high_m    = numeric()
    ))
  }
  nn_idx <- st_nearest_feature(pts, high)
  tibble(
    !!id_col := pts[[id_col]],
    nearest_high_parcel_index = high$parcel_index[nn_idx],
    dist_to_nearest_high_m    = as.numeric(
      st_distance(pts, high[nn_idx, ], by_element = TRUE)
    )
  )
}

mits_nearhigh <- nearest_high(mits, mit_id_col, high_parcels)
apps_nearhigh <- nearest_high(apps, app_id_col, high_parcels)

mits_direct <- mits_direct |> left_join(mits_nearhigh, by = mit_id_col)
apps_direct <- apps_direct |> left_join(apps_nearhigh, by = app_id_col)
```

---

## 5  Write outputs

```{r write-outputs}
st_write(mits_direct, file.path(out_dir, "alt_mits_direct.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
st_write(apps_direct, file.path(out_dir, "alt_apps_direct.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

mits_direct |> st_drop_geometry() |>
  write_csv(file.path(out_dir, "alt_mits_direct_review.csv"))
apps_direct |> st_drop_geometry() |>
  write_csv(file.path(out_dir, "alt_apps_direct_review.csv"))
```

---

## Summary

### Match status counts

Row counts reflect the fan-out: a single input record landing on *N* parcel
polygons contributes *N* rows here.

```{r summary-counts}
bind_rows(
  mits_direct |> st_drop_geometry() |> mutate(side = "mitigations"),
  apps_direct |> st_drop_geometry() |> mutate(side = "applications")
) |>
  count(side, match_status) |>
  pivot_wider(names_from = match_status, values_from = n, values_fill = 0L) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = "Row counts by side and match status (rows = input record × parcel hit)"
  )
```

### Shared-footprint fan-out

How often does a single input record intersect multiple parcel polygons?
Condo complexes and stacked parcels show up as `n_parcels >= 2`.

```{r multihit-diagnostic}
multihit <- function(df, id_col, label) {
  df |>
    st_drop_geometry() |>
    filter(!is.na(parcel_index)) |>
    count(.data[[id_col]], name = "n_parcels") |>
    count(n_parcels, name = "n_records") |>
    mutate(side = label)
}

bind_rows(
  multihit(mits_direct, mit_id_col, "mitigations"),
  multihit(apps_direct, app_id_col, "applications")
) |>
  pivot_wider(names_from = side, values_from = n_records, values_fill = 0L) |>
  arrange(n_parcels) |>
  knitr::kable(
    format.args = list(big.mark = ","),
    caption = "Distribution of parcel-hit counts per input record (n_parcels > 1 = shared-footprint)"
  )
```

### Distance to nearest high-priority parcel (low-priority / off-parcel only)

If most low-hit records are within ~30 m of a high-priority parcel the low
classification is likely a geocode artefact, not a meaningful difference in
parcel type.

```{r proximity-distribution}
bind_rows(
  mits_direct |> st_drop_geometry() |> mutate(side = "mitigations"),
  apps_direct |> st_drop_geometry() |> mutate(side = "applications")
) |>
  filter(match_status %in% c("matched_low", "no_parcel_hit")) |>
  group_by(side, match_status) |>
  summarise(
    n      = n(),
    p10    = round(quantile(dist_to_nearest_high_m, 0.10, na.rm = TRUE), 1),
    median = round(median(dist_to_nearest_high_m,          na.rm = TRUE), 1),
    p90    = round(quantile(dist_to_nearest_high_m, 0.90, na.rm = TRUE), 1),
    max    = round(max(dist_to_nearest_high_m,             na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  knitr::kable(
    caption = "Distance (m) from low-priority / no-hit records to nearest high-priority parcel"
  )
```

---

## Review browsers

Records that landed on a low-priority LUC or off-parcel entirely, sorted
ascending by distance to the nearest high-priority parcel. Rows near the top
are the most suspicious (close to a high-priority parcel but not on one).

### Mitigations

```{r review-browser-mits}
#| column: page

mits_direct |>
  st_drop_geometry() |>
  filter(match_status %in% c("matched_low", "no_parcel_hit")) |>
  arrange(dist_to_nearest_high_m) |>
  DT::datatable(
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = "Mitigations on low-priority LUCs or off-parcel, sorted by distance to nearest high-priority parcel."
  )
```

### Applications

```{r review-browser-apps}
#| column: page

apps_direct |>
  st_drop_geometry() |>
  filter(match_status %in% c("matched_low", "no_parcel_hit")) |>
  arrange(dist_to_nearest_high_m) |>
  DT::datatable(
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 25, scrollX = TRUE),
    caption  = "Applications on low-priority LUCs or off-parcel, sorted by distance to nearest high-priority parcel."
  )
```

---

## Output schema

Each row of `alt_mits_direct.gpkg` / `alt_apps_direct.gpkg`:

| Column | Source | Notes |
|---|---|---|
| *(all original mit/app columns)* | input | unchanged |
| `parcel_index` … `ASSESSED.TOTAL.VALUE` | parcels | NA when `no_parcel_hit` |
| `luc_tier_num` | derived | `as.integer(luc_tier)` |
| `match_status` | derived | `matched_high` / `matched_low` / `no_parcel_hit` |
| `nearest_high_parcel_index` | derived | `parcel_index` of closest high-priority parcel |
| `dist_to_nearest_high_m` | derived | metres; review only, not used for matching |

Same columns in the CSVs, without geometry.

---

## Unresolved — confirm before using outputs

- **`HIGH_TIER_MIN`** is set to `3L` (tiers 3–4: vacant, gov, all residential).
  Change to `4L` for tier-4-only (1–4 family + site address).
- **`mit_id_col` / `app_id_col`** default to `"unique_id"`. If the input files
  don't carry that column, substitute the appropriate HMA/IHP primary key.
