---
title: "Step 5: Mitigations to Parcels"
author: "Russell Blessing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show / hide code"
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This notebook spatially matches mitigation and application records to parcels
using a tiered priority scheme. Parcels are scored and ranked as high, medium,
or low priority for matching based on land use code and address quality.
For each priority tier, mitigation/application geometries are first matched via
direct intersection with parcel polygons, then by nearest-neighbor distance
(≤ 50 m) for those that do not intersect. Unmatched records after all three
tiers fall back to a global nearest-neighbor match across all parcels.

**Input files (Russell's):**

- `nc1_cl_merge.csv` — parcel–CoreLogic merged table (Step 4 output)
- `parcels_study_area.gpkg` — OneMap parcel geometries with `geometry_id`
- `mit_newgeo.gpkg` — mitigations (re-geocoded)
- `mit_elevation_qaqc_filled.csv` — elevation QAQC scores for mitigations
- `applications_final.gpkg` — FEMA applications
- `NC_huc6_data_union_sf.shp` — flood study area boundary

**External files (written to `out_dir`):**

- `parcels_matcheable.gpkg` — filtered parcels before priority scoring
- `parcels_pri.gpkg` — parcels with priority scores
- `mit_highpri.csv`, `mit_medpri.csv`, `mit_lowpri.csv`, `mit_dist.csv`
- `app_highpri.csv`, `app_medpri.csv`, `app_lowpri.csv`, `app_dist.csv`
- `app_sa_full_noqa.gpkg`, `mit_sa_full_noqa.gpkg`
- `apps_pcls.gpkg`, `mits_pcls.gpkg`

```{r libraries}
library(sf)
library(dplyr)
library(readr)
library(stringr)
library(purrr)
library(tidyr)
```

```{r setup}
out_dir  <- "/proj/mhinolab/users/rbless/data/Obstacles_Output"

# External data paths
flood_boundary_path <- "/proj/mhinolab/projects/obstacles/NC_huc6_data_union_sf/NC_huc6_data_union_sf.shp"
mit_path            <- "/proj/mhinolab/projects/obstacles2/CRC_May_2025/Raw Data Inputs/raw_app_ncb/mit_newgeo.gpkg"
mit_elev_path       <- "/proj/mhinolab/projects/obstacles2/CRC_May_2025/Raw Data Inputs/raw_app_ncb/mit_elevation_qaqc_filled.csv"
app_path            <- "/proj/mhinolab/projects/obstacles2/CRC_May_2025/Raw Data Inputs/raw_app_ncb/applications_final.gpkg"
geoms_path          <- file.path(out_dir, "parcels_study_area.gpkg")
```

---

## 1  Load & prepare parcels

```{r load-parcels}
#| eval: false

# load-parcels — remove the mit_sa/app_sa mutates entirely
parcels_raw <- read_csv(
  file.path(out_dir, "nc1_cl_merge.csv"),
  col_types = cols(.default = col_character())
) |>
  mutate(parcel_index = as.integer(parcel_index))

nc1_geoms <- st_read(geoms_path, quiet = TRUE) |>
  mutate(
    parcel_index = as.integer(parcel_index),
    geometry_id  = as.integer(factor(as.character(geom)))
  ) |>
  select(parcel_index, geometry_id)

parcels <- parcels_raw |>
  select(-any_of(c("geometry_id1", "index_right", "geometry_id", "geometry"))) |>
  left_join(st_drop_geometry(nc1_geoms), by = "parcel_index") |>
  left_join(nc1_geoms |> select(parcel_index, geometry = geom), by = "parcel_index") |>
  st_as_sf(crs = st_crs(nc1_geoms))
```

---

## 2  Filter invalid parcels

```{r filter-parcels}
#| eval: false

# Remove water parcels (flagged via parcel number keywords)
water_keywords <- "bay|creek|sound|water|ocean|river|canal|pond|inlet"

parcels <- parcels |>
  filter(!str_detect(parno, regex(water_keywords, ignore_case = TRUE)))

# Flag and remove invalid parcels
parcels <- parcels |>
  mutate(
    invalidparcels = case_when(
      # No identifying info at all
      is.na(`LAND USE CODE`) & parno == "" & is.na(siteadd) & mailadd == "" ~ 1L,
      # Carteret: no land use code, blank parno → likely right-of-way
      is.na(`LAND USE CODE`) & parno == "" & cntyfips == "031" ~ 1L,
      # Brunswick: no parno or altparno → right-of-way
      is.na(`LAND USE CODE`) & parno == "" & altparno == "" & cntyfips == "019" ~ 1L,
      # Explicit ROW tags (excluding "ROWAN COUNTY")
      (str_detect(parno,    regex("row", ignore_case = TRUE)) |
       str_detect(altparno, regex("row", ignore_case = TRUE))) &
        parno != "ROWAN COUNTY" ~ 1L,
      TRUE ~ 0L
    )
  ) |>
  filter(invalidparcels == 0L)

# Save pre-priority filtered parcels for diagnostic mapping
st_write(parcels |> select(-invalidparcels),
         file.path(out_dir, "parcels_matcheable.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

parcels <- parcels |> select(-invalidparcels)
```

---

## 3  Parcel priority scoring

Land use codes that define each priority bucket:

```{r luc-definitions}
luc_vacant      <- c("465","460","455","454","453","452","450","400")
luc_gov         <- c("604","603","601")           # muni/county/tax-exempt owned
luc_residential <- c("165","163","160","151","138","137","136","135","134","133",
                      "132","131","117","116","115","113","112","111","109","106",
                      "103","102","100")
luc_fr_1_4      <- c("165","163","160","151","138","137","136","135","132","115",
                      "102","100")
luc_top         <- c(luc_vacant, luc_gov, luc_residential)
luc_bottom      <- c("605","606","607","609","610","614","615","618","620","630",
                      "800","818","839","850","860","872","875","877","880","883",
                      "885","886","114")
```

> **Scoring change from original Python script**
>
> The original script collapsed vacant, government-owned, and all residential
> parcels into a single `qual_luc = 3` bucket, meaning a 1–4 family home and
> a municipal lot were treated identically. It also weighted site address and
> mailing address equally (1 pt each), and the composite `qual_sum` only
> ranged 0–5, making the high/med/low thresholds somewhat arbitrary.
>
> The revised scheme:
>
> | LUC group | `luc_tier` | Rationale |
> |---|---|---|
> | 1–4 family residential (`luc_fr_1_4`) | 4 | Primary buyout/mitigation candidate |
> | Vacant land (`luc_vacant`) | 3 | High candidate, no structure risk |
> | Gov-owned (`luc_gov`) | 3 | Eligible but less common |
> | Other residential (`luc_residential \ luc_fr_1_4`) | 2 | Lower priority residential |
> | Unknown LUC | 1 | Insufficient info |
> | Commercial/industrial (`luc_bottom`) | 0 | Unlikely candidates |
>
> Site address is weighted **2 pts** (more useful for geocoding/matching than
> mailing address), mailing address **1 pt**, giving a `qual_sum` range of
> 0–7 and more granular priority tiers.
>
> The `gov` list also had a missing comma bug in the Python
> (`['604' '603', '601']` → string `'604603'`), so no parcel ever matched it.
> Fixed here.

```{r score-parcels}
#| eval: false

parcels <- parcels |>
  mutate(
    `LAND USE CODE` = as.character(`LAND USE CODE`),

    # Land use flags (kept for downstream use)
    vacant  = if_else(`LAND USE CODE` %in% luc_vacant,      1L, 0L),
    gov_lu  = if_else(`LAND USE CODE` %in% luc_gov,         1L, 0L),
    res_lu  = if_else(`LAND USE CODE` %in% luc_residential, 1L, 0L),
    fr_1_4  = if_else(`LAND USE CODE` %in% luc_fr_1_4,      1L, 0L),

    # Address quality: site address weighted 2x, mail address 1x
    qual_siteadd = if_else(!is.na(siteadd) & siteadd != "", 2L, 0L),
    qual_mailadd = if_else(!is.na(mailadd) & mailadd != "", 1L, 0L),

    # Tiered LUC score (0–4)
    luc_tier = case_when(
      `LAND USE CODE` %in% luc_fr_1_4                                        ~ 4L,
      `LAND USE CODE` %in% luc_vacant                                         ~ 3L,
      `LAND USE CODE` %in% luc_gov                                            ~ 3L,
      `LAND USE CODE` %in% luc_residential & !`LAND USE CODE` %in% luc_fr_1_4 ~ 2L,
      `LAND USE CODE` %in% luc_bottom                                         ~ 0L,
      TRUE                                                                     ~ 1L
    ),

    # Composite score (0–7) and priority tier
    qual_sum = luc_tier + qual_siteadd + qual_mailadd,
    match_priority = case_when(
      luc_tier == 4 & qual_siteadd == 2 ~ "high",   # 1-4 family + site address
      luc_tier >= 3 | qual_sum >= 5     ~ "med",
      TRUE                              ~ "low"
    )
  )

# Save prioritised parcel layer
st_write(parcels, file.path(out_dir, "parcels_pri.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
```

```{r priority-split}
#| eval: false

# Split into priority tiers
highpri <- parcels |> filter(match_priority == "high")
medpri  <- parcels |> filter(match_priority == "med")
lowpri  <- parcels |> filter(match_priority == "low")
```

---

## 4  Load & prepare mitigations and applications

```{r load-flood-boundary}
flood_boundary <- st_read(flood_boundary_path, quiet = TRUE) |>
  st_transform(crs = 32119)
```

```{r load-mit}
mit_gdf <- st_read(mit_path, quiet = TRUE) |>
  st_transform(crs = st_crs(flood_boundary))

# Merge elevation QAQC scores
mit_elev <- read_csv(mit_elev_path, col_types = cols(.default = col_character())) |>
  rename_with(str_trim)

mit_gdf <- mit_gdf |>
  left_join(mit_elev |> 
  select(record_ID, elevation_score), by = "record_ID")

# Clip to study area
mit_gdf <- mit_gdf |>
  mutate(Inside_SA = as.integer(lengths(st_within(st_geometry(mit_gdf), flood_boundary)) > 0))

mit_sa <- mit_gdf |> filter(Inside_SA == 1)
```

```{r load-app}
app <- st_read(app_path, quiet = TRUE) |>
  st_transform(crs = st_crs(flood_boundary))

app <- app |>
  mutate(
    app_id    = row_number(),
    Inside_SA = as.integer(lengths(st_within(st_geometry(app), flood_boundary)) > 0)
  )

app_sa <- app |> filter(Inside_SA == 1)
```

```{r unique-geoms-setup}
mit_geom_col <- attr(mit_sa, "sf_column")
app_geom_col <- attr(app_sa, "sf_column")

mit_sa <- mit_sa |>
  mutate(g_indicator = as.integer(factor(as.character(st_geometry(mit_sa)))))

app_sa <- app_sa |>
  mutate(
    g_indicator = as.integer(factor(as.character(st_geometry(app_sa)))),
    app_id      = row_number()
  )

unique_mit <- mit_sa |> distinct(g_indicator, .keep_all = TRUE) |>
  select(g_indicator, all_of(mit_geom_col)) |>
  st_transform(crs = st_crs(flood_boundary))

unique_app <- app_sa |> distinct(g_indicator, .keep_all = TRUE) |>
  select(g_indicator, all_of(app_geom_col)) |>
  st_transform(crs = st_crs(flood_boundary))
```

---

## 5  Build unique-geometry sets

Matching is done on unique geometries then merged back, avoiding redundant
spatial operations where multiple records share a coordinate.

```{r load-parcels-priority}
read_priority <- function(priority, path) {
  st_read(
    path,
    query = paste0("SELECT match_priority, geom FROM \"parcels_pri\" WHERE match_priority = '", priority, "'"),
    quiet = TRUE
  ) |>
    st_transform(crs = 32119)
}

parcels_path <- file.path(out_dir, "parcels_pri.gpkg")

highpri <- read_priority("high", parcels_path)
medpri  <- read_priority("med",  parcels_path)
lowpri  <- read_priority("low",  parcels_path)
gc()
```


```{r unique-geoms-parcels}
# Reproject parcel priority sets to EPSG:32119
to_unique_geoms <- function(df) {
  geom_col <- attr(df, "sf_column")
  df |>
    st_transform(crs = 32119) |>
    (\(x) mutate(x, geometry_id = as.integer(factor(as.character(st_geometry(x))))))() |>
    distinct(geometry_id, .keep_all = TRUE) |>
    select(geometry_id, all_of(geom_col)) |>
    rename(geometry = all_of(geom_col))
}

ug_high <- to_unique_geoms(highpri)
ug_med  <- to_unique_geoms(medpri)  |> filter(!geometry_id %in% ug_high$geometry_id)
ug_low  <- to_unique_geoms(lowpri)  |> filter(!geometry_id %in% ug_high$geometry_id,
                                               !geometry_id %in% ug_med$geometry_id)
```


---

## 6  Tiered spatial matching helper

A single function runs the intersect → distance (≤ 50 m) workflow for one
priority tier, for both mitigations and applications.

```{r match-function}
#' Match unique geometries to a parcel tier via intersect + nearest-neighbour
#'
#' @param candidates sf with columns g_indicator + geometry (mit or app unique geoms)
#' @param parcel_tier sf with columns geometry_id + geometry (unique parcel geoms)
#' @param max_dist   maximum allowable distance in metres for distance matches
#'
#' @return tibble with columns g_indicator, geometry_id, and optionally distance
match_to_tier <- function(candidates, parcel_tier, max_dist = 50) {

  # --- intersect pass ---
  intersect_raw <- st_join(candidates, parcel_tier, join = st_intersects, left = TRUE)

  intersect_matched <- intersect_raw |>
    filter(!is.na(geometry_id)) |>
    st_drop_geometry() |>
    mutate(intersect = 1L) |>
    distinct(g_indicator, .keep_all = TRUE)

  unmatched <- intersect_raw |>
    filter(is.na(geometry_id)) |>
    select(g_indicator)

  # --- distance pass on the remainder ---
  if (nrow(unmatched) == 0 || nrow(parcel_tier) == 0) {
    dist_matched <- tibble(g_indicator = integer(), geometry_id = integer(),
                           distance = numeric())
  } else {
    dist_joined <- st_join(unmatched, parcel_tier, join = st_nearest_feature)

    parcel_geoms <- parcel_tier$geometry[match(dist_joined$geometry_id, parcel_tier$geometry_id)]

    dist_matched <- dist_joined |>
      mutate(distance = as.numeric(
        st_distance(st_geometry(dist_joined), parcel_geoms, by_element = TRUE)
      )) |>
      st_drop_geometry() |>
      distinct(g_indicator, .keep_all = TRUE) |>
      filter(distance <= max_dist)
  }

  bind_rows(
    intersect_matched |> select(g_indicator, geometry_id),
    dist_matched      |> select(g_indicator, geometry_id, any_of("distance"))
  )
}
```

---

## 7  Run tiered matching

```{r match-high}
mit_highpri <- match_to_tier(unique_mit, ug_high)
app_highpri <- match_to_tier(unique_app, ug_high)

# Remove matched records before next tier
mits_rem <- unique_mit |> filter(!g_indicator %in% mit_highpri$g_indicator)
apps_rem <- unique_app |> filter(!g_indicator %in% app_highpri$g_indicator)
```

```{r match-med}
mit_medpri <- match_to_tier(mits_rem, ug_med)
app_medpri <- match_to_tier(apps_rem, ug_med)

mits_rem <- mits_rem |> filter(!g_indicator %in% mit_medpri$g_indicator)
apps_rem <- apps_rem |> filter(!g_indicator %in% app_medpri$g_indicator)
```

```{r match-low}
mit_lowpri <- match_to_tier(mits_rem, ug_low)
app_lowpri <- match_to_tier(apps_rem, ug_low)

mits_rem <- mits_rem |> filter(!g_indicator %in% mit_lowpri$g_indicator)
apps_rem <- apps_rem |> filter(!g_indicator %in% app_lowpri$g_indicator)
```

```{r match-fallback}
# Fallback: global nearest-neighbour across all parcel tiers (no distance cap)
ug_all <- bind_rows(ug_high, ug_med, ug_low)

nn_fallback <- function(remaining, all_geoms) {
  if (nrow(remaining) == 0) {
    return(tibble(g_indicator = integer(), geometry_id = integer(),
                  distance = numeric()))
  }
  joined <- st_join(remaining, all_geoms, join = st_nearest_feature)
  joined |>
    mutate(distance = as.numeric(
      st_distance(
        st_geometry(joined),
        all_geoms$geometry[match(geometry_id, all_geoms$geometry_id)],
        by_element = TRUE
      )
    )) |>
    st_drop_geometry() |>
    distinct(g_indicator, .keep_all = TRUE) |>
    select(g_indicator, geometry_id, distance)
}

mit_dist <- nn_fallback(mits_rem, ug_all)
app_dist <- nn_fallback(apps_rem, ug_all)
```

---

## 8  Save per-tier CSVs

```{r save-tier-csvs}
write_csv(mit_highpri, file.path(out_dir, "mit_highpri.csv"))
write_csv(app_highpri, file.path(out_dir, "app_highpri.csv"))
write_csv(mit_medpri,  file.path(out_dir, "mit_medpri.csv"))
write_csv(app_medpri,  file.path(out_dir, "app_medpri.csv"))
write_csv(mit_lowpri,  file.path(out_dir, "mit_lowpri.csv"))
write_csv(app_lowpri,  file.path(out_dir, "app_lowpri.csv"))
write_csv(mit_dist,    file.path(out_dir, "mit_dist.csv"))
write_csv(app_dist,    file.path(out_dir, "app_dist.csv"))
```

---

## 9  Merge geometry tags back to full mit / app datasets

```{r merge-back}
mit_geom_full <- bind_rows(mit_highpri, mit_medpri, mit_lowpri, mit_dist)
app_geom_full <- bind_rows(app_highpri, app_medpri, app_lowpri, app_dist)

mit_sa_full <- mit_sa |>
  mutate(unique_id = row_number()) |>
  left_join(mit_geom_full, by = "g_indicator")

app_sa_full <- app_sa |>
  mutate(unique_id = row_number()) |>
  left_join(app_geom_full, by = "g_indicator")

# Save geometry-tagged full datasets before QA
st_write(app_sa_full, file.path(out_dir, "app_sa_full_noqa.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
st_write(mit_sa_full, file.path(out_dir, "mit_sa_full_noqa.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
```

---

## 10  Join back to parcels

```{r join-to-parcels}
parcels <- st_read(parcels_path, quiet = TRUE)

# Reconstruct nc1_geoms (needed for final geometry join; load-parcels is eval:false)
nc1_geoms <- st_read(geoms_path, quiet = TRUE) |>
  mutate(
    parcel_index = as.integer(parcel_index),
    geometry_id  = as.integer(factor(as.character(geom)))
  ) |>
  select(parcel_index, geometry_id, geom)

# Drop geometry from app/mit before joining to avoid geometry column conflicts
apps_nogeom <- app_sa_full |> st_drop_geometry()
mits_nogeom <- mit_sa_full |> st_drop_geometry()

parcels_nogeom <- parcels |>
  select(-any_of(c("vacant","gov_lu","res_lu","fr_1_4",
                    "qual_siteadd","qual_mailadd","qual_luc",
                    "qual_sum","match_priority")))

apps_pcls <- apps_nogeom |>
  left_join(st_drop_geometry(parcels_nogeom), by = "geometry_id") |>
  distinct() |>
  left_join(nc1_geoms |> select(parcel_index, geometry = geom), by = "parcel_index") |>
  st_as_sf(crs = st_crs(nc1_geoms))

mits_pcls <- mits_nogeom |>
  left_join(st_drop_geometry(parcels_nogeom), by = "geometry_id") |>
  distinct() |>
  left_join(nc1_geoms |> select(parcel_index, geometry = geom), by = "parcel_index") |>
  st_as_sf(crs = st_crs(nc1_geoms))

st_write(apps_pcls, file.path(out_dir, "apps_pcls.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
st_write(mits_pcls, file.path(out_dir, "mits_pcls.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)
```

---

## Summary

```{r summary}
tibble::tibble(
  Stage = c(
    "Mitigations in study area",
    "Applications in study area",
    "Mit matched — high priority",
    "App matched — high priority",
    "Mit matched — med priority",
    "App matched — med priority",
    "Mit matched — low priority",
    "App matched — low priority",
    "Mit fallback (global NN)",
    "App fallback (global NN)",
    "Mit unmatched",
    "App unmatched"
  ),
  N = c(
    nrow(mit_sa),
    nrow(app_sa),
    nrow(mit_highpri),
    nrow(app_highpri),
    nrow(mit_medpri),
    nrow(app_medpri),
    nrow(mit_lowpri),
    nrow(app_lowpri),
    nrow(mit_dist),
    nrow(app_dist),
    nrow(mits_rem),
    nrow(apps_rem)
  )
)
```
