---
title: "Step 8: Manuscript Figures"
author: "Russell Blessing"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show / hide code"
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This notebook builds the manuscript figures from the finalized outputs of the
earlier pipeline steps. The first deliverable is **Fig. 2** — the
*flooded-to-funded* pipeline funnel.

The pipeline tracks residential properties through five sequential, nested
stages:

| # | Stage | Definition | Source |
|---|-------|------------|--------|
| 1 | **flooded** | intersected the footprint of ≥ 1 of the 78 modelled events | Step 7 |
| 2 | **eligible** | flooded during an event for which its county received a PDD | Step 7 |
| 3 | **in a community** | its community (city, else county) is one where the local government assisted with the submission of ≥ 1 HMA application | Step 5 `apps_pcls` |
| 4 | **applied** | the property itself is named on an HMA application | Step 5 `apps_pcls` |
| 5 | **funded** | the property received a mitigation | Step 5 `mits_pcls` |


The funnel is computed twice: once for **all** flood-exposed residential
properties, and once for the subset exposed to flooding **more than once**.



```{r libraries}
library(sf)
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(ggplot2)
library(ggridges)
```

---

## 1  Configuration

```{r config}
out_dir <- "/proj/mhinolab/users/rbless/data/Obstacles_Output"

cfg <- list(
  master_path      = file.path(out_dir, "parcels_master.gpkg"),
  repeat_threshold = 2L
)
```

---

## 2  Diagnostic — what is actually in the files

```{r diagnostic}
needed <- c("parcel_index", "cntyfips", "SITUS.CITY",
            "flooded_any", "eligible", "fr_1_4", "interpolate",
            "value", "cent_lat", "cent_lng", "eif_lat", "eif_lon",
            "applied", "funded", "pct_white_2020", "pct_black_2020")

master_cols <- names(st_read(cfg$master_path, quiet = TRUE,
  query = "SELECT * FROM \"parcels_master\" LIMIT 0"))

for (w in needed) {
  cat(if (w %in% master_cols) "[ok]  " else "[MISS] ", w, "\n", sep = "")
}
```

---

## 3  Load finalized datasets

```{r load-parcels}
parcels <- st_read(cfg$master_path, quiet = TRUE) |>
  st_drop_geometry() |>
  transmute(
    parcel_index   = as.integer(parcel_index),
    county         = as.character(cntyfips),
    city           = as.character(SITUS.CITY),
    flood_ct       = suppressWarnings(as.numeric(flooded_any)),
    elig_ct        = suppressWarnings(as.numeric(eligible)),
    res_flag       = suppressWarnings(as.integer(fr_1_4)),
    interpolate    = suppressWarnings(as.integer(interpolate)),
    value          = suppressWarnings(as.numeric(value)),
    lat            = suppressWarnings(as.numeric(cent_lat)),
    lng            = suppressWarnings(as.numeric(cent_lng)),
    eif_lat        = suppressWarnings(as.numeric(eif_lat)),
    eif_lon        = suppressWarnings(as.numeric(eif_lon)),
    applied        = as.logical(applied),
    funded         = as.logical(funded),
    pct_white_2020 = suppressWarnings(as.numeric(pct_white_2020)),
    pct_black_2020 = suppressWarnings(as.numeric(pct_black_2020))
  ) |>
  mutate(
    community = if_else(is.na(city) | str_trim(city) == "",
                        paste0("COUNTY_", county),
                        str_to_upper(str_trim(city)))
  )

cat("parcels loaded:", format(nrow(parcels), big.mark = ","), "\n")
```

---

## 4  Build per-parcel pipeline flags

```{r build-flags}
# Set of communities where the local government assisted with >=1 HMA
# application (draft Fig. 2 stage-3 definition).
app_communities <- parcels |>
  filter(applied) |>
  distinct(community) |>
  pull(community)

cat("communities with >=1 application:", length(app_communities), "\n")

pipeline <- parcels |>
  # restrict to residential properties (Fig. 2 universe)
  filter(res_flag == 1L | interpolate == 1L) |>
  mutate(
    s1_flooded   = !is.na(flood_ct) & flood_ct >= 1,
    s2_eligible  = !is.na(elig_ct)  & elig_ct  >= 1,
    s3_community = community %in% app_communities,
    s4_applied   = applied,
    s5_funded    = funded,
    repeat_exp   = !is.na(flood_ct) & flood_ct >= cfg$repeat_threshold
  ) |>
  # cumulative "reached stage k" — enforces nesting
  mutate(
    reach1 = s1_flooded,
    reach2 = reach1 & s2_eligible,
    reach3 = reach2 & s3_community,
    reach4 = reach3 & s4_applied,
    reach5 = reach4 & s5_funded
  )
```

---

## 5  Funnel table

```{r funnel-table}
funnel_for <- function(df, label) {
  r <- c(sum(df$reach1), sum(df$reach2), sum(df$reach3),
         sum(df$reach4), sum(df$reach5))
  tibble(
    group        = label,
    stage_num    = 1:5,
    stage        = factor(
      c("Flooded", "Eligible", "In a community",
        "Applied", "Funded"),
      levels = c("Flooded", "Eligible", "In a community",
                 "Applied", "Funded")),
    n            = r,
    pct_of_prev  = c(NA, r[-1] / r[-5]),         # stage-to-stage retention
    pct_of_flood = r / r[1]                       # cumulative share of flooded
  )
}

funnel_all    <- funnel_for(pipeline,                     "All flood-exposed")
funnel_repeat <- funnel_for(filter(pipeline, repeat_exp), "Repeat flood-exposed")

funnel <- bind_rows(funnel_all, funnel_repeat)

funnel |>
  mutate(
    n            = format(n, big.mark = ","),
    `% of prev`  = if_else(is.na(pct_of_prev), "—",
                           paste0(round(pct_of_prev * 100, 1), "%")),
    `% of flood` = paste0(round(pct_of_flood * 100, 1), "%")
  ) |>
  select(group, stage, n, `% of prev`, `% of flood`) |>
  knitr::kable()
```

---

## 7  Fig. 3 — property value distribution by pipeline stage
 
Each property's value is converted to a **percentile** against the entire
`fr_1_4` study-area sample. Histograms use 5% bins; the red vertical line
marks the stage median. Four stages shown (Eligible dropped).
 
### 7.1  Build the percentile sample
 
```{r fig3-build-sample}
# Attach value + lat/lng to the residential pipeline (fr_1_4) parcels.
# Drop parcels with missing or non-positive value and missing coordinates.
value_df <- pipeline |>
  filter(!is.na(value), value > 0, !is.na(lat), !is.na(lng)) |>
  mutate(
    eif_cell = paste0(eif_lat, "_", eif_lon),
    pct_A = dplyr::percent_rank(value) * 100
  ) |>
  group_by(eif_cell) |>
  mutate(
    cell_n = dplyr::n(),
    pct_B  = if_else(cell_n >= 5, dplyr::percent_rank(value) * 100, NA_real_)
  ) |>
  ungroup()

cat("sample parcels (fr_1_4 OR interp, w/ value + coords):",
    format(nrow(value_df), big.mark = ","), "\n")
cat("distinct EIF cells:",
    format(dplyr::n_distinct(value_df$eif_cell), big.mark = ","), "\n")
cat("parcels with valid Panel B percentile:",
    format(sum(!is.na(value_df$pct_B)), big.mark = ","), "\n")
```
 
### 7.2  Plot

```{r fig3-hist, fig.width=6, fig.height=7}
stage_levels_4 <- c("Flooded", "In a community", "Applied", "Funded")

fig3_hist_data <- bind_rows(
  value_df |> filter(reach1) |> transmute(stage = "Flooded",        pct = pct_A),
  value_df |> filter(reach3) |> transmute(stage = "In a community", pct = pct_A),
  value_df |> filter(reach4) |> transmute(stage = "Applied",        pct = pct_A),
  value_df |> filter(reach5) |> transmute(stage = "Funded",         pct = pct_A)
) |>
  mutate(stage = factor(stage, levels = stage_levels_4))

fig3_medians <- fig3_hist_data |>
  group_by(stage) |>
  summarise(median_pct = median(pct, na.rm = TRUE), .groups = "drop")

fig3 <- ggplot(fig3_hist_data, aes(x = pct)) +
  geom_histogram(binwidth = 5, boundary = 0,
                 fill = "grey60", colour = "grey30", linewidth = 0.2) +
  geom_vline(data = fig3_medians,
             aes(xintercept = median_pct),
             colour = "firebrick", linewidth = 0.6) +
  facet_wrap(~ stage, ncol = 1, scales = "free_y", strip.position = "right") +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%"),
                     expand = c(0.005, 0)) +
  labs(
    title = "Fig. 3  Property value percentile distribution by HMA stage",
    x = "Property value percentile (study-area sample)",
    y = "Number of parcels"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text.y     = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.spacing.y  = unit(0.5, "lines"),
    plot.title       = element_text(face = "bold")
  )

fig3
```

```{r fig3-save}
dir.create(file.path(getwd(), "figures"), showWarnings = FALSE)
ggsave(file.path(getwd(), "figures", "fig3.png"),
       fig3, width = 6, height = 7, dpi = 300)
write_csv(fig3_medians, file.path(getwd(), "figures", "fig3_medians.csv"))
```
---

## 8  Fig. 4 — neighborhood racial composition by HMA stage

Two-panel figure following the draft's Fig. 4 caption.
**Panel A** is % Non-Hispanic White residents in the EIF 0.01° grid cell
containing each property. **Panel B** is % Black residents in the same cell.
A cell is at stage *k* if it contains ≥ 1 residential parcel that reached *k*.
Cell-level. Four stages shown (Eligible dropped); 5% bins; red line = median.

### 8.1  Build per-panel data

```{r fig4-hist, fig.width=8, fig.height=9}
# Per-cell race shares — one row per EIF cell, with stage flags.
# Multiply by 100 if `pct_white_2020` / `pct_black_2020` on master are 0-1
# fractions; remove the *100 if they are already 0-100 percentages.
cells_at_stage <- value_df |>
  group_by(eif_cell) |>
  summarise(
    pct_white = dplyr::first(pct_white_2020) * 100,
    pct_black = dplyr::first(pct_black_2020) * 100,
    s1 = any(reach1, na.rm = TRUE),
    s3 = any(reach3, na.rm = TRUE),
    s4 = any(reach4, na.rm = TRUE),
    s5 = any(reach5, na.rm = TRUE),
    .groups = "drop"
  )

fig4_hist_data <- cells_at_stage |>
  tidyr::pivot_longer(c(s1, s3, s4, s5),
                      names_to = "stage_key", values_to = "in_stage") |>
  filter(in_stage) |>
  mutate(stage = recode(stage_key,
    s1 = "Flooded", s3 = "In a community",
    s4 = "Applied", s5 = "Funded"
  )) |>
  select(stage, pct_white, pct_black) |>
  tidyr::pivot_longer(c(pct_white, pct_black),
                      names_to = "race", values_to = "pct") |>
  mutate(
    stage = factor(stage, levels = stage_levels_4),
    race  = factor(recode(race,
      pct_white = "A. % Non-Hispanic White",
      pct_black = "B. % Black"
    ), levels = c("A. % Non-Hispanic White", "B. % Black"))
  )

fig4_medians <- fig4_hist_data |>
  group_by(stage, race) |>
  summarise(median_pct = median(pct, na.rm = TRUE), .groups = "drop")

fig4 <- ggplot(fig4_hist_data, aes(x = pct)) +
  geom_histogram(binwidth = 5, boundary = 0,
                 fill = "grey60", colour = "grey30", linewidth = 0.2) +
  geom_vline(data = fig4_medians,
             aes(xintercept = median_pct),
             colour = "firebrick", linewidth = 0.6) +
  facet_grid(stage ~ race, scales = "free_y", switch = "y") +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%"),
                     expand = c(0.005, 0)) +
  labs(
    title = "Fig. 4  Neighborhood racial composition by HMA stage",
    x = "Percent of residents (EIF cell)",
    y = "Number of cells"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text       = element_text(face = "bold"),
    strip.placement  = "outside",
    panel.grid.minor = element_blank(),
    panel.spacing    = unit(0.5, "lines"),
    plot.title       = element_text(face = "bold")
  )

fig4
```

### 8.2  Medians table

```{r fig4-medians}
fig4_medians |>
  mutate(
    n          = NA_integer_,   # counts live in fig4_hist_data if needed
    median_pct = round(median_pct, 1)
  ) |>
  select(race, stage, median_pct) |>
  arrange(race, stage) |>
  knitr::kable(col.names = c("Panel", "Stage", "Median pct"))
```

```{r fig4-save}
dir.create(file.path(getwd(), "figures"), showWarnings = FALSE)
ggsave(file.path(getwd(), "figures", "fig4.png"),
       fig4, width = 8, height = 9, dpi = 300)
write_csv(fig4_medians, file.path(getwd(), "figures", "fig4_medians.csv"))
```